Automatic Matching of Bullet Lands

### 2015-12-30 imaging-paper.Rnw
took out table to save some space

\section{Introduction}
\hh{
Todo list for journal shape-up: there is a lot of good material, we mainly need to focus on what story to tell and make that a solid argument.
\begin{enumerate}
%\item take out a few things: no circular fit, no figures 9 and 10 
%\item do we need the web app?
%\item out: lots of the figures can go into a supplement, together with a discussion. In the paper we then have a summary of what can be seen, so that we can focus on the results. 
%\item add: formula for loess fit and its s.e.
%\item add: comparison of error rates to those of FEs in Hamby study.
%\item change: emphasis in varying cross section. once a stable region is found, how little change is happening? 
%\item add: Why do we take the lowest stable cross section?
%\item add: more discussion of the density plots: probability ratio for a match - exchanged probability ratios for ROC curves
\item add: more discussion on additional bullet images. NIST database claims to have 70 bullets -- that would be two sets of the study, but only 35 bullets are currently available. Should we contact Alan Zheng to see whether we can get access to more bullet files?
\item add: should we add a paragraph on the speed of the evaluation? - separate out into pre-processing (needs to be done once) and matching (pairwise comparison of lands). If we only do bullet matching, we could speed this process up by eliminating some land comparisons.
\item add: we might need a proper definition of the joint striation marks (the result of striation-identify)
\item change: we could make the table into a labelled dotplot of importance to save space. We don't need the mean and standard deviations anymore -- the density plots and the ROCs take care of that.
\item change: the focus of the paper is initially quite strongly on the maximal consecutively matching striae (CMS). We need to emphasis the role of CMS in the introduction: max CMS is Biasotti's main criterium for matching.  max CMS is part of current practice (AFTE criteria citation). AFIS \cite{AFIS} is suggesting it for an automated system (was there ever a phase II of that project?). Problem is, that determining CMS is subjective; counting CMS in particular, has been found to be subjective to the examiner: `the concept of counting striations is subjective and based on experience'~\citep{miller:1998}.  For an automatic routine we are faced with the problem of having to determine what a stria is. XXX Should we include one of those microscopic images to bring that point home? We would need somebody qualified to tell us how many CMS there are, though.
\item to consider: AOAS states that accepted papers rarely have more than 20 pages. We are currently at about 23--24 pages.
\item definitions: 
A bullet's {\it signature} is defined as the residual surface measurements after removing the big structure using the loess fit. The {\it cross section} defines the location along the $x$ axis  at which the signature is taken.
\end{enumerate}
}
Firearm examination is a forensic tool used to determine whether two bullets were fired from the same gun barrel. This process has broad applicability in terms of  convictions  in the United States criminal justice system. Ballistics identification has long been considered an accepted and reliable procedure, but in the past ten years has undergone more significant scrutiny. In 2005, in \emph{United States vs. Green}, the court ruled that the forensic expert could not confirm that the bullet casings came from a specific weapon with certainty, but could merely ``describe" other casings which are similar. Further court cases in the late 2000s expressed caution about the use of ballistics identification evidence~\citep{giannelli:2011}.

@@

The primary goal in this type of forensic analysis is to identify these striation markings in an attempt to match two bullets. Since gun barrels should produce similar markings on bullets fired consecutively, the careful analysis of these surface images can help us determine whether two bullets are a match. However, this process has traditionally required forensic experts to perform the analysis, and required physical examination of the bullets. We propose a framework which allows for the analysis of the surface topologies, and the transcription of the markings into a 2D plotting framework which we believe makes identification of matches easier and more rooted in science for both the expert and the lay person.

\hh{XXX More detailed discussion of automatic approaches and database discussion XXX}

@@

\begin{figure}[hbtp]
  \centering
<<tree:forest, echo=FALSE, fig.width=5, fig.height=5.25, out.width='.5\\textwidth', warning = FALSE>>=
qplot(pred, forest, geom="jitter", data=bstats, colour=match, 
      shape=match) + theme_bw() +
  theme(legend.position="bottom") +
  geom_vline(xintercept=0.5, colour="grey50") +
  geom_hline(yintercept=0.5, colour="grey50") + 
  xlab("Predictions from decision tree") +
  ylab("Predictions from random forest")   +
  scale_colour_brewer(palette="Paired", labels=c("KNM", "KM")) +
  scale_shape_discrete(labels=c("KNM", "KM"))
@
\caption{\label{fig:tree-forest}Scatterplot of prediction results from the tree (horizontal) and the forest (vertical). Each point corresponds to a land-to-land comparison. Known matches are indicated in blue. Matches predicted by the decision tree are on the right half of the plot, matches predicted by the random forest are in the upper half. }
\end{figure}

Besides a probabilistic quantification of matches, we get from a random forest an assessment of the importance of each of the features derived from the bullets' 3d topological surface measurements. Table~\ref{tab:importance} shows an overview of the importance of each variable measured as the mean decrease in the gini index when the variable in question is included in a tree.
\begin{table}[tbhp]
\caption{\label{tab:importance}Table of  features derived from bullet image ordered by importance in predicting matches. Importance is measured in terms of mean decrease in gini index when including the variable in a decision tree. Averages (and standard deviations) for known matches (KM) and known non-matches (KNM) are shown in the last four columns.}
\centering
<<importance, echo=FALSE, results='asis', warning=FALSE>>=
imp <- data.frame(importance(rtrees))
imp$Variable <- row.names(imp)
imp <- imp[order(-imp$MeanDecreaseGini),]
names(imp)[1] <- "Importance"
imp <- imp[,2:1]
imp$meanMatch <- sapply(imp$Variable, function(var) {
  mean(subset(bstats, match)[,var])
})
imp$sdMatch <- sapply(imp$Variable, function(var) {
  sd(subset(bstats, match)[,var])
})
imp$meanNonMatch <- sapply(imp$Variable, function(var) {
  mean(subset(bstats, !match)[,var])
})
imp$sdNonMatch <- sapply(imp$Variable, function(var) {
  sd(subset(bstats, !match)[,var])
})
imp$meanMatch <- round(imp$meanMatch,1)
imp$sdMatch <- sprintf("(%5.2f)", imp$sdMatch)
imp$meanNonMatch <- round(imp$meanNonMatch,1)
imp$sdNonMatch <- sprintf("(%7.2f)", imp$sdNonMatch)
row.names(imp) <- 1:nrow(imp)
imp$Variable[c(2,4)] <- c("#matches", "#non-matches")
names(imp)[-(1:2)] <- c("KM", "(sd)", "KNM", "(sd)")
print(xtable(imp, align="clrrd{3.2}rd{3.2}", digits=c(0,0,1,1,2,1,2)), 
      floating = FALSE)
@
\end{table}

The rows in the table are ordered according to descending importance. The variable with the most predictive power by far is cross-correlation, followed by the number of matches and non-matches. CMS is found only in sixth place.

@@

By suggesting an automated algorithm that first removes class characteristics, such as the grooves and the curvature of the bullet to reveal the region of the  land, then identifies peaks and valleys on this land, we reduce subjectivity and with it possible sources of bias. Counting CMS in particular, has been found to be subjective to the examiner: `the concept of counting striations is subjective and based on experience'~\citep{miller:1998}.

@@

Only a much larger database and transparent communication of results will give us  answers to whether automatic matching of ballistic evidence is really feasible. Whether there really is individualization of this evidence, as is often silently assumed. And in the case, that individualization is a reasonable assumption, whether  there is a way to be able to distinguish between individual and subclass characteristics.

>>>>

\section{Introduction}
\hh{
Todo list for journal shape-up: there is a lot of good material, we mainly need to focus on what story to tell and make that a solid argument.
\begin{enumerate}
%\item take out a few things: no circular fit, no figures 9 and 10 
%\item do we need the web app?
%\item out: lots of the figures can go into a supplement, together with a discussion. In the paper we then have a summary of what can be seen, so that we can focus on the results. 
%\item add: formula for loess fit and its s.e.
%\item add: comparison of error rates to those of FEs in Hamby study.
%\item change: emphasis in varying cross section. once a stable region is found, how little change is happening? 
%\item add: Why do we take the lowest stable cross section?
%\item add: more discussion of the density plots: probability ratio for a match - exchanged probability ratios for ROC curves
\item add: more discussion on additional bullet images. NIST database claims to have 70 bullets -- that would be two sets of the study, but only 35 bullets are currently available. Should we contact Alan Zheng to see whether we can get access to more bullet files?
\item add: we could extend the results of the James Hamby study by trying to match every single land against all other lands. This would almost double the number of comparisons ($21,115 = (118+88)\cdot 205/2$). Something for the appendix? What would be fun, would be to show all of the aligned signatures for all ten barrels.
\item add: should we add a paragraph on the speed of the evaluation? - separate out into pre-processing (needs to be done once) and matching (pairwise comparison of lands). If we only do bullet matching, we could speed this process up by eliminating some land comparisons.
\item add: we might need a proper definition of the joint striation marks (the result of striation-identify)
\item change: we could make the table into a labelled dotplot of importance to save space. We don't need the mean and standard deviations anymore -- the density plots and the ROCs take care of that.
%\item change: the focus of the paper is initially quite strongly on the maximal consecutively matching striae (CMS). We need to emphasis the role of CMS in the introduction: max CMS is Biasotti's main criterium for matching.  max CMS is part of current practice (AFTE criteria citation). AFIS \cite{AFIS} is suggesting it for an automated system (was there ever a phase II of that project?). Problem is, that determining CMS is subjective; counting CMS in particular, has been found to be subjective to the examiner: `the concept of counting striations is subjective and based on experience'~\citep{miller:1998}.  For an automatic routine we are faced with the problem of having to determine what a stria is. XXX Should we include one of those microscopic images to bring that point home? We would need somebody qualified to tell us how many CMS there are, though.
\item to consider: AOAS states that accepted papers rarely have more than 20 pages. We are currently over by a couple of pages.
\item definitions: 
A bullet's {\it signature} is defined as the residual surface measurements after removing the big structure using the loess fit. The {\it cross section} defines the location along the $x$ axis  at which the signature is taken.
\end{enumerate}
}
Firearm examination is a forensic tool used to determine whether two bullets were fired from the same gun barrel. This process has broad applicability in terms of  convictions  in the United States criminal justice system. Ballistics identification has long been considered an accepted and reliable procedure, but in the past ten years has undergone more significant scrutiny. In 2005, in \emph{United States vs. Green}, the court ruled that the forensic expert could not confirm that the bullet casings came from a specific weapon with certainty, but could merely ``describe" other casings which are similar. Further court cases in the late 2000s expressed caution about the use of ballistics identification evidence~\citep{giannelli:2011}.

@@

The primary goal in this type of forensic analysis is to identify these striation markings in an attempt to match two bullets. Since gun barrels should produce similar markings on bullets fired consecutively, the careful analysis of these surface images can help in determining whether two bullets are a match. This process has traditionally required forensic experts to perform the analysis, and required physical examination of the bullets.

\hh{
Current standard practice~\cite{afte:1992} relies in part on the assessment of the so-called maximal number of consecutively matching striae (CMS), first defined by \citet{biasotti:1959}.
One of the problems is that a human inspection to determine CMS is subjective \citep{miller:1998}. 
}

In this paper, we propose a framework which allows for the automatic analysis of the surface topologies of bullets, and the transcription of the individual characteristics into a 2D plotting framework. 
This allows for an objective and quantitative assessment of striae-based bullet matches.

%\hh{XXX More detailed discussion of automatic approaches and database discussion XXX}

@@

\begin{figure}[hbtp]
  \centering
<<tree:forest, echo=FALSE, fig.width=5, fig.height=2.5, out.width='.6\\textwidth', warning = FALSE>>=
predm <- melt(bstats, measure.var=c("pred", "forest"))
levels(predm$variable) <- c("Tree", "Forest")
predm$match <- c("KNM", "KM")[as.numeric(predm$match)+1]
qplot(data=predm, value, variable, geom="jitter", colour=match, shape=match) + theme_bw() + facet_grid(match~., labeller="label_both") +
    scale_colour_brewer(palette="Paired", labels=c("KNM", "KM")) +
  scale_shape_discrete(labels=c("KNM", "KM")) + ylab("") + 
  xlab("Predicted probability of a match") +
  theme(legend.position="none")

# qplot(pred, forest, geom="jitter", data=bstats, colour=match, 
#       shape=match) + theme_bw() +
#   theme(legend.position="bottom") +
#   geom_vline(xintercept=0.5, colour="grey50") +
#   geom_hline(yintercept=0.5, colour="grey50") + 
#   xlab("Predictions from decision tree") +
#   ylab("Predictions from random forest")   +
#   scale_colour_brewer(palette="Paired", labels=c("KNM", "KM")) +
#   scale_shape_discrete(labels=c("KNM", "KM"))
# 
@
\caption{\label{fig:tree-forest}Prediction results from the tree and the forest. Using a cut-off probability of 0.5 the forest correctly predicts every single comparison. Compared to the tree, the forest's prediction probabilities are  shrunk towards either end of the prediction range. }
\end{figure}

Besides a probabilistic quantification of matches, we get from a random forest an assessment of the importance of each of the features derived from the bullets' 3d topological surface measurements. Figure~\ref{fig:importance} shows an overview of the importance of each variable measured as the mean decrease in the gini index when the variable in question is included in a tree.
\begin{figure}
<<echo=FALSE, fig.width=7, fig.height=1.5, out.width='.7\\textwidth'>>=
imp <- data.frame(importance(rtrees))
imp$Variable <- row.names(imp)
imp$Variable[c(7,8)] <- c("#matches", "#non-matches")
imp1 <- subset(imp, Variable != c("x1", "sd.D"))
imp2 <- subset(imp, Variable %in% c("x1", "sd.D"))
qplot(MeanDecreaseGini,1, data=imp, geom="point") + 
  geom_text(aes(y=1.1, label=Variable), angle=45, hjust=0, data=imp1) + 
  geom_text(aes(y=0.9, label=Variable), angle=-45, hjust=0, data=imp2) + 
  ylim(c(0.5, 2.5)) + theme_bw() + ylab("") +
  theme(axis.ticks.y=element_blank(), axis.text.y=element_blank(),
        panel.grid.major.y=element_blank(), panel.grid.minor.y=element_blank()) +
  xlab("Importance (mean decrease in Gini index)")
@
\caption{\label{fig:importance}Importance of features in the random forest. Importance is measured in terms of mean decrease in gini index when including the variable in a decision tree.}
\end{figure}
\begin{table}[tbhp]
\caption{\label{tab:importance}Table of  features derived from bullet image ordered by importance in predicting matches. Importance is measured in terms of mean decrease in gini index when including the variable in a decision tree. Averages (and standard deviations) for known matches (KM) and known non-matches (KNM) are shown in the last four columns.}
\centering
<<importance, echo=FALSE, results='asis', warning=FALSE>>=
imp <- data.frame(importance(rtrees))
imp$Variable <- row.names(imp)
imp <- imp[order(-imp$MeanDecreaseGini),]
names(imp)[1] <- "Importance"
imp <- imp[,2:1]
imp$meanMatch <- sapply(imp$Variable, function(var) {
  mean(subset(bstats, match)[,var])
})
imp$sdMatch <- sapply(imp$Variable, function(var) {
  sd(subset(bstats, match)[,var])
})
imp$meanNonMatch <- sapply(imp$Variable, function(var) {
  mean(subset(bstats, !match)[,var])
})
imp$sdNonMatch <- sapply(imp$Variable, function(var) {
  sd(subset(bstats, !match)[,var])
})
imp$meanMatch <- round(imp$meanMatch,1)
imp$sdMatch <- sprintf("(%5.2f)", imp$sdMatch)
imp$meanNonMatch <- round(imp$meanNonMatch,1)
imp$sdNonMatch <- sprintf("(%7.2f)", imp$sdNonMatch)
row.names(imp) <- 1:nrow(imp)
imp$Variable[c(2,4)] <- c("#matches", "#non-matches")
names(imp)[-(1:2)] <- c("KM", "(sd)", "KNM", "(sd)")
print(xtable(imp, align="clrrd{3.2}rd{3.2}", digits=c(0,0,1,1,2,1,2)), 
      floating = FALSE)
@
\end{table}

%The rows in the table are ordered according to descending importance. 
The variables with the most predictive power by far are cross-correlation and the overall number of matches, followed by the total depth of joint striations $S$ and non-matches. CMS is found only in sixth place.

@@

By suggesting an automated algorithm that first removes class characteristics, such as the grooves and the curvature of the bullet to reveal the region of the  land, then identifies peaks and valleys on this land, we reduce subjectivity and with it possible sources of bias. In particular, `the concept of counting striations is subjective and based on experience'~\citep{miller:1998}.

@@

Only a much larger database and transparent communication of results will give us  answers to whether automatic matching of ballistic evidence is really feasible,  whether there really is individualization of this evidence, as is often silently assumed. And in the case, that individualization is a reasonable assumption, whether  there is a way to be able to distinguish between individual and subclass characteristics.
